\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex
Tweaks

Occasional evolutionary leaps, in which a large number of mutations
are acquired in sudden jumps,
is a signature feature of
SARS-CoV-2~\citep{Corey2021-sars,otto2021origins,Nielsen2023-host}.
Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
current HMM parameterization of three mutations per recombination (mismatch
ratio). The first few sequences from these new lineages will be a poor match to
the existing ARG, and the HMM will therefore
search for ways to reduce the number of mutations required by recombining
segments.
This appears to be the case for the Delta VoC,
corresponding to rows A and B of Table~\ref{tab:false_positive},
which emerged and quickly rose to high prevalence in early-mid 2021,
carrying 30 characteristic mutations compared to the reference sequence
\citep{McCrone2022-context}, including nine mutations in the S gene
not seen in earlier VoCs.
The origins of the  Delta VoC are
are illustrated in  Figure~\ref{fig:false_positives},
which shows the large numbers of mutations involved,
and the multiple closely related recombinations inferred
before tree-like behaviour is resumed at the ancestor of
[FIXME what percentage?] Delta samples (node \texttt{tsk261771}).
Similar patterns around the emergence of Delta are seen in the Wide ARG
(not shown).

@@

\begin{figure} \centering
\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
\caption{\label{fig:false_positives}
Subgraph of the Long ARG, focusing on two likely false positive
recombination nodes at the start of the Delta wave
(``focal'' nodes, in red, corresponding to rows A and B in
Table~\ref{tab:false_positive}).
The path to node \texttt{tsk261771} has been expanded:
this node (in gold) is the ancestor of ${\sim}89.8\%$ of Delta samples in the Long ARG,
and represents a large polytomy with 107 immediate children. Of the remaining Delta samples, 10\% of Delta
samples are mostly descendants of the node \texttt{tsk232088} on the far right,
a sibling of the earliest focal recombination node.
The suspected-incorrect recombination path to an early B.1 MRCA is also shown.
Note the large amount of additional recombination (black nodes) among close
descendants of the focal nodes.
% Why say this here and not at for the other subgraph figs?
%See supplementary file \protect\path{Subgraph_sample_mapping.txt} to  map tskID to EPI\_ISL and strain name.
}
\end{figure}

>>>>

Occasional evolutionary leaps, in which a large number of mutations
are acquired in sudden jumps,
is a signature feature of
SARS-CoV-2~\citep{Corey2021-sars,otto2021origins,Nielsen2023-host}.
Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
current HMM parameterization of three mutations per recombination (mismatch
ratio). The first few sequences from these new lineages will be a poor match to
the existing ARG, and the HMM will therefore
search for ways to reduce the number of mutations required by recombining
segments.
This appears to be the case for the Delta VoC,
corresponding to rows A and B of Table~\ref{tab:false_positive},
which emerged and quickly rose to high prevalence in early-mid 2021,
carrying 30 characteristic mutations compared to the reference sequence
\citep{McCrone2022-context}, including nine mutations in the S gene
not seen in earlier VoCs.
The origins of the  Delta VoC are
are illustrated in  Figure~\ref{fig:false_positives},
which shows the large numbers of mutations involved,
and the multiple closely related recombinations inferred
before tree-like behaviour is resumed at the ancestor of
90\% of Delta samples (node \texttt{tsk261771}).
Similar patterns around the emergence of Delta are seen in the Wide ARG
(not shown).

@@

\begin{figure} \centering
\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
\caption{\label{fig:false_positives}
Subgraph of the Long ARG, focusing on two likely false positive
recombination nodes at the start of the Delta wave
(``focal'' nodes, in red, corresponding to rows A and B in
Table~\ref{tab:false_positive}).
The path to node \texttt{tsk261771} has been expanded:
this node (in gold) is the ancestor of ${\sim}89.8\%$ of Delta samples in the Long ARG,
and represents a large polytomy with 107 immediate children. Of the remaining Delta samples, most (8.4\%)
are descendants of the node \texttt{tsk232088} on the far right,
a sibling of the earliest focal recombination node.
The suspected-incorrect recombination path to an early B.1 MRCA is also shown.
Note the large amount of additional recombination (black nodes) among close
descendants of the focal nodes.
% Why say this here and not at for the other subgraph figs?
%See supplementary file \protect\path{Subgraph_sample_mapping.txt} to  map tskID to EPI\_ISL and strain name.
}
\end{figure}
